---
title: "MB5370 Module 2 Workshop 4"
author: "Caitlin Buckle"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  warning = FALSE
)
```

This is a Quarto document for Module 2 Workshop 4 exercises.

#### Loading Packages

```{r}
#All packages used in this file can be loaded from here 
#Installed sf, terra, and tmap in the console for this workshop 
library(tidyverse)
library(sf)
library(terra)
library(tmap)
```


#### THE PROBLEM: An academic has given you an opportunity to collaborate on a new dataset she is working with - it has extensive samples of copepod richness throughout Australian oceans and the Southern Ocean too. She has a hypothesis - that copepod species richness will be higher in warmer waters than cooler waters, but she needs help sorting the data. She wants you to use your skills in R to help develop a map that could help you visualise whether this hypothesis is worth pursuing. 

### Section 4.6: Downloading and Loading the Spatial Dataset 

```{r}
data<-read_csv("../Module_2_Data_Science/data/workshop4/copepods_raw.csv")
data #5313 observations, 11 variables 
#silk_id is the ID for each silk, and each silk has been broken into segment numbers (segment_no) for processing ease. 
```

### Section 4.7: Data Exploration 

First step to making our map using ggplot2 is to plot the coordinates for the samples.

```{r}
ggplot(data)+
  aes(x=longitude,y=latitude,colour=richness_raw)+
  geom_point() #NOTE: this is not a map, simply a scatter plot but still a nice and easy way to visualise your spatial data 
```

Can also check for other patterns, such as in our richness data 

```{r}
ggplot(data)+
  aes(x=latitude,y=richness_raw)+
  stat_smooth()+
  geom_point()
#unusual change in data pattern at latitude -44. Why?? 
```

### Section 4.8: Getting Going with Maps 

To start mapping, we need to turn our point data into a spatially reference data frame using the sf() package (an open standard for geospatial databases). The equivalent of a shape file or feature collection in GIS. 

```{r}
sdata<-st_as_sf(data,coords=c("longitude","latitude"),
                crs=5126)
#st_as_sf converts different data types to simple features, coords gives the names of the columns that relate to spatial coordinates (in order of X coordinate followed by Y coordinate), crs us coordinate reference system 
```

### Section 4.9: Coordinate Reference Systems 

```{r}
crs5126<-st_crs(5126)
crs5126 #look at the whole CRS
crs5126$Name #pull out just the name of the CRS 
crs5126$wkt
```

