---
title: "MB5370 Module Two Workshop Two Keystone Exercise"
author: "Caitlin Buckle"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

The Premise: You have inherited a project from a former PhD student who was studying the relationship between water quality and predatory fish assemblages along the Ross River Estuary gradient. Unfortunately, their data management was entirely manual, and the files are a disaster.

They left you with four separate datasets:
    estuary_catch_log.xlsx: Handwritten field counts of fish species (using common names), spread across multiple spreadsheet tabs.
    estuary_metadata.csv: The spatial coordinates and estuary zones for each site.
    estuary_sonde_data.csv: High-frequency water quality sensor readings.
    species_dictionary.csv: A taxonomic lookup table linking local common names to their accepted scientific names.

Your mission is to write a single, reproducible R script (or .qmd) that ingests this mess, cleans it, joins it into a single master dataset, and produces a publication-ready visualisation of the estuary's ecology.

Note: the datasets are now available in the class repository

## PHASE ONE: Ingestion and Decontamination 

Load the four datasets and think about how to clean and wrangle them individually 

```{r}
#Load packages 
library(tidyverse) #includes dplyr, ggplot2, lubridate
library(readxl) #for reading excel files
```

#### Load files into R 
```{r}
#Load .csv files using read_csv (files are stored in workshop2 folder inside data folder)
metadata<- read_csv("../Module_2_Data_Science/data/workshop2/estuary_metadata.csv")

sonde_data<- read_csv("../Module_2_Data_Science/data/workshop2/estuary_sonde_data.csv")

species_dict<-read.csv("../Module_2_Data_Science/data/workshop2/species_dictionary.csv")

catch_log_1<- read_excel("../Module_2_Data_Science/data/workshop2/estuary_catch_log.xlsx", sheet = "Lower_ross")
catch_log_2<- read_excel("../Module_2_Data_Science/data/workshop2/estuary_catch_log.xlsx", sheet = "Mid_Ross ")
catch_log_3<- read_excel("../Module_2_Data_Science/data/workshop2/estuary_catch_log.xlsx", sheet = "ROSS MOUTH")
catch_log_4<- read_excel("../Module_2_Data_Science/data/workshop2/estuary_catch_log.xlsx", sheet = "Upper Ross")
catch_log<-bind_rows(catch_log_1, catch_log_2, catch_log_3, catch_log_4) #individually loaded in each sheet from excel file, so had to bind them into one data frame --> catch_log
```

### Standardise all site and species columns to lowercase with underscores 

```{r}
clean_catch<-catch_log%>% 
  mutate(
    site_clean=str_to_lower(site), #convert to lowercase
    site_clean=str_trim(site_clean), #remove whitespace
    site_clean=str_replace_all(site_clean,pattern=" ",replacement="_") #replace with underscores where needed
  )%>% 
  mutate(
    species_clean=str_to_lower(species), #convert to lowercase
    species_clean=str_trim(species_clean), #remove whitespace
    species_clean=str_replace_all(species_clean,pattern=" ",replacement="_") #replace spaces between species names with underscores
  )
print(clean_catch) #much better! 

clean_metadata<-metadata%>%
  mutate(
    site_clean=str_to_lower(site_name),
    site_clean=str_trim(site_clean), 
    site_clean=str_replace_all(site_clean,pattern=" ",replacement="_")
  )
print(clean_metadata)
```

### Fix temporal parsing in sonde data 

```{r}
clean_sonde<-sonde_data%>%
  mutate(
     true_time=dmy_hm(timestamp) #telling mutate that our preferred time structure is day-month-year-hour-minute
  )
print(clean_sonde) #new column called true_time with format of day-month-year-hour-minute 
```

### Remove -999 values from turbidity sensors 

```{r}
# Convert the -999 error codes to true NA values
cleaned_sonde<- clean_sonde%>%
  mutate(turbidity= na_if(turbidity, -999)) 

print(cleaned_sonde)
summary(cleaned_sonde) #all -999 changed into NA, now have 567 NA's in the data set 
```

## PHASE TWO: THE RELATIONAL ARCHITECTURE 

Bringing the data together into a cohesive formula 

```{r}
#using floor_date() inside labridate to group temperature values for each site on each day 
sonde_daily<-cleaned_sonde%>% 
  mutate( 
    day=floor_date(true_time,unit="day") #rounds the timestamps down to just the date
    )%>%
  group_by(site,day)%>% #group by the site and your new daily date column 
  summarise(
    mean_temp=mean(temperature,na.rm=TRUE), 
    mean_salinity=mean(salinity,na.rm=TRUE),
    .groups="drop"
  )
print(sonde_daily) #tibble has 120 rows and four variables, grouped by day and then a column for mean_temp and mean_salinity 
```

### Taxonomic translation 

Need to merge species_dictionary into your catch data, replacing the common names with accurate scientific taxonomy.

```{r}
cc_scientific<-clean_catch %>%
  left_join(species_dict,by=join_by(species_clean==common_name)) #joining species_dict to clean_catch, have specified that species_clean in clean_watch is equal to common_name in species_dict
  

print(cc_scientific)
```

### Merge translated catch date, spatial metadata and your daily water quality summaries into a single master data frame. 

Translated catch data = cc_scientific 
Spatial metadata = clean_metadata
Daily water quality summaries = sonde_daily 

```{r}
master <- cc_scientific %>%
  left_join(clean_metadata, by = join_by(site_clean == site_clean)) %>%  #join clean_metadata to cc_scientific
  mutate(date_fixed = ymd(date)) %>% #fix date column in new data frame (specify it is ymd)
  left_join(sonde_daily, by = join_by(site_clean == site, date_fixed == day)) #join sonde_daily to other joins

print(master)

master_df <- master %>%
  select(-c(site, site_name, species, date_fixed)) #remove unnecessary columns 

print(master_df)
```


## PHASE THREE: ZERO CATCH FRAMEWORK 

The student only recorded a row when they actually caught a fish, meaning there is no zeros in the data frame (which are important ecologically). 

```{r}
final_master_matrix <- master_df %>%
  complete(site_clean, date, scientific_name) %>% #build the zero-data framework across sites, dates, and species
  mutate(count = coalesce(count, 0)) %>% #replace NA values in the count column with 0
  group_by(site_clean, date) %>% 
  fill(lat, lon, zone, mean_temp, mean_salinity, .direction = "downup") %>% #carry over day-specific environment data (lat, long, temp, salinity)
  ungroup()

print(final_master_matrix)
```
## PHASE FOUR: STATISTICAL EXTRACTION 

From your master data set, generate a clean summary table showing the mean ± standard error of both the fish catch and the salinity for each scientific species across each estuary zone.

```{r}
summary_table <- final_master_matrix %>%
 group_by(zone, scientific_name) %>%  #group by your target categories
  summarise( #calculate mean and standard error for both variables
    mean_catch = mean(count, na.rm = TRUE),
    se_catch   = sd(count, na.rm = TRUE) / sqrt(n()), #mean fish catch and se calculation
    zone_mean_salinity = mean(mean_salinity, na.rm = TRUE),
    zone_se_salinity   = sd(mean_salinity, na.rm = TRUE) / sqrt(n()),
    .groups = "drop" 
  )

print(summary_table)
```

## PHASE FIVE: VISUAL COMMUNICATION

Prove that your data rescue was successful. Construct a multi-panel plot using ggplot2 and facet_wrap().
The Goal: Visualise how the abundance of different species changes along the salinity gradient.
The Requirement: The plot should have professional axis labels, a clean theme (e.g., theme_minimal()), be ordered spatially (e.g. from upstream to downstream), and the facet labels (the scientific names) should really be italicised!

```{r}
#Order four zones spatially (from low salinity to high)
plot_data <- final_master_matrix %>%
  mutate(zone = factor(zone, levels = c("Upstream", "Middle", "Downstream", "Marine")))

#Build the multi-panel plot
ggplot(plot_data, aes(x = mean_salinity, y = count, color = zone)) +
  geom_jitter(alpha = 0.4, width = 0.5, height = 0.2) + #add jittered points so overlapping zeros at the bottom become visible
  geom_smooth(method = "loess", se = TRUE, color = "black", size = 0.8, linetype = "dashed") + # add a smoothed trend line to show the abundance trend across increasing salinity
  facet_wrap(~ scientific_name, scales = "fixed") + # Create a panel for each species 
  theme_minimal(base_size = 13) +
  labs(
    x = "Mean Salinity (ppt)", #x-axis title 
    y = "Species Abundance (Count)", #y-axis title 
    title = "Species Abundance Along An Estuarine Salinity Gradient", #plot title 
    color = "Estuary Zone" # colour by estuary zone (using levels as defined before)
  ) +
  theme(
    strip.text = element_text(face = "italic", size = 11, color = "black"),  #using element_text(face = "italic") to italicise scientific names
    legend.position = "right", #legend on the right of the plot 
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40")
  )
```


